---
title: "Storm Data: Population Health and Economic Damages"
author: "Lindsay Spencer"
date: "April 10, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

# Synopsis

Storms can be costly in a number of different ways. This research focuses on their cost to public health and the cost to the economy. The goal of this research is to find which storm events are the most costly in both areas. We find that tornados, excessive heat, and lightning have the highest impact to population health and floods and hurricanes have the highest economic costs. 

# Data Processing

### The technical information

This research was performed on a CPU running Windows 10 Pro, with Intel(R) Core(TM) i7-4650U CPU @ 1.7GH and 2.3GHz.

Using R Studio Version 0.99.893 and R version 3.2.4 Revised.

The file was last downloaded on April 10, 2016 at 6:30 PM PST. 

## Load Data

First, we ensure that all the packages we will need are loaded and ready to process our data. 

```{r loadPackages, hide = TRUE}
list.of.packages <- c("lubridate", "sqldf", "ggplot2", "gridExtra")
new.packages <-
    list.of.packages[!(list.of.packages %in% installed.packages()
                       [, "Package"])]

if (length(new.packages))
    install.packages(new.packages)

library(lubridate)
library(sqldf)
library(ggplot2)
library(gridExtra)
```

Next, we download the storm data from the course website. 

``` {r downloadFile}
fileUrl <-
     "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
fileName = "stormData.csv"

if (!file.exists(fileName)) {
     download.file(fileUrl, fileName)
}
```

And we read in the file to the stormActivityData variable. 

``` {r readData}
stormActivityData <- read.csv(fileName)
```

## Data Cleaning

### Subseting the Relevant Data 

Let's take a look at the raw data. 

``` {r stormActivityData, out.width='\\textwidth'}
head(stormActivityData)
```

From this, we know we are only interested in a subset of the actual data. Specifically, we want the event type (EVTYPE), the number of fatalities (FATALITIES), the number of injuries (INJURIES), the property damage amount (PROPDMG and PROPDMGEXP), and the crop damage amount (CROPDMG and CROPDMGEXP). We will also use the date (BGN_DATE) to remove some of the data for dates that are probably incomplete.

``` {r subsetData}
relevantData <- c(
    "BGN_DATE",
    "EVTYPE",
    "FATALITIES",
    "INJURIES",
    "PROPDMG",
    "PROPDMGEXP",
    "CROPDMG",
    "CROPDMGEXP"
)

relevantDataSubset <- stormActivityData[relevantData]
```

Let's take a look at this subsetted data.

``` {r relevantDataSubset}
head(relevantDataSubset)
```

This is good, but we still need to do some transformations to be able to use this data for our analysis. 

### Caclulating Total Damages 

The amount of damages to property and crops are presented as a number and an exponential factor. The documentation indicates that they use "K" for thousands, "M" for millions, and "B" for billions. The function, getMultiplier, was written to translate the EXP variable into a multiplier to calculate the correct amount of damages. It uses k and K, m and M, b and B, and ignores any other symbol that may be in the EXP column.

``` {r getMultiplier}
getMultiplier <- function(multiplier) {
     ifelse(
          multiplier == 'K' || multiplier == 'k',
          mult <- 1000,
          ifelse(
               multiplier == 'M' || multiplier == 'm',
               mult <- 1000000,
               ifelse(
                    multiplier == 'B' ||
                         multiplier == 'b',
                    mult <- 1000000000,
                    mult <- 1
               )
          )
     )
     mult
}
```

This function is now used to correctly record the amount of property damage and the amount of crop damage each event caused.

``` {r damageCalculations}
relevantDataSubset$propDamage <- relevantDataSubset$PROPDMG *
    as.integer(lapply(relevantDataSubset$PROPDMGEXP, getMultiplier))

relevantDataSubset$cropDamage <- relevantDataSubset$CROPDMG *
    as.integer(lapply(relevantDataSubset$CROPDMGEXP, getMultiplier))
```

Using these amounts, we calculate the total damage caused by each event. 

``` {r totalDamageCalculation}
relevantDataSubset$totalDamage <-
    relevantDataSubset$propDamage + relevantDataSubset$cropDamage
```

Here's a look at what our dataset looks like after our calculations. 

``` {r relevantDataSubsetPart2}
head(relevantDataSubset)
```

### Subseting Years that Have More Recorded Events

From the documentation, we know that the earlier years probably do not contain complete information. A look at the histogram of number of events by year shows us that this is probably true. To do this, we'll have to extract the year from the date column.

``` {r histogram}
relevantDataSubset$YEAR <-
    year(as.Date(relevantDataSubset$BGN_DATE, '%m/%d/%Y'))

hist(relevantDataSubset$YEAR, xlab = "Year", 
     main = "Number of Events per Year", col = "green")
```

This shows us clearly that the earlier years do not have nearly the number of events recorded as the later years do. We make the assumption that the number of events recorded is at fault and not that the number of events has been increasing, so we will only look at the second half of our data. To do this, we will calculate the 50th percentile and subset the data to only look at events from the more recent years.

``` {r newestData}
cutOffYear <- quantile(relevantDataSubset$YEAR, c(.5))
cutOffYear
```

We will use this value in the next section to exclude data from the first half of the given timeframe.

### Aggregate Totals by Event

Now, we will total the amount of fatalities, injuries, and total damage done by each type of event. We will also use this opportunity to exclude the first 50% of the data and to rename the headers. 

``` {r sumByEvent, hide=TRUE}
selectStatement <- paste(
    "SELECT ",
    "EVTYPE AS Event,",
    "SUM(FATALITIES) AS Fatalities,",
    "SUM(INJURIES) AS Injuries,",
    "SUM(TotalDamage) AS Damage",
    "FROM relevantDataSubset ",
    "WHERE YEAR >= '",
    cutOffYear,
    "'",
    "GROUP BY EVTYPE"
)

finalDataSet <- sqldf(selectStatement)
```

Here is a sample of the final dataset.

``` {r finalDataSet}
head(finalDataSet)
```

# Results

## Question 1

#### Across the United States, which types of events are most harmful with respect to population health?

To answer this question, we will plot both the top 5 fatal events and top 5 injury-producing events separately.

``` {r populationHealth}
topFatalEvents <-
    head(finalDataSet[order(-finalDataSet$Fatalities), ], 5)

g1 <-
    ggplot(topFatalEvents, 
           aes(x = factor(Event, levels = unique(Event)), y = Fatalities)) +
    geom_bar(stat = "identity", fill = "blue") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x = "Event")

topInjuryEvents <-
    head(finalDataSet[order(-finalDataSet$Injuries), ], 5)

g2 <-
    ggplot(topInjuryEvents, 
           aes(x = factor(Event, levels = unique(Event)), y = Injuries)) +
    geom_bar(stat = "identity", fill = "red") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x = "Event")

grid.arrange(g1, g2, ncol = 2)
```

We can see that tornados cause both the greatest amount of fatalities and injuries, and that excessive heat causes the second most in both categories. Lightning is in the top 5 for both, as well. We can conclude that tornados, excessive heat, and lightning are the most harmful events with respect to public health. 

## Question 2

#### Across the United States, which types of events have the greatest economic consequences?

For this question, we'll plot the top 5 events with the greatest total economic damage. 

``` {r economicDamage}
economicDamageEvents <-
    head(finalDataSet[order(-finalDataSet$Damage),], 5)

ggplot(economicDamageEvents,
       aes(x = factor(Event, levels = unique(Event)), y = Damage)) +
    geom_bar(stat = "identity", fill = "orange") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x = "Event", y = "$ Damage (in Billions)") +
    scale_y_continuous(
        labels = function(n) {
            format(n / 1000000000, scientific = FALSE)
        }
    )
```

From this graph, we see that floods cause the greatest economic damage, with hurricanes and typhoons coming in second. 

# Conclusion

Our analysis has shown us that tornados, excessive heat, and lightning are the most harmful events with respect to public health; while floods, hurricanes, and typhoons have the highest economic consequences. 



